---
title: "Functional principal component analysis for spatial summary functions"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Multiplex FPCA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  figs.out = "../figures"
)
invisible(suppressPackageStartupMessages(library(tidyverse)))
```

The `{mxfda}` package contains tools for analyzing single-cell multiplex imaging data using methods from functional data analysis. Analyses for this package are executed and stored using an S4 object of class `mxFDA`. This vignette outlines how to perform feature extraction using functional principal component analysis and multilevel functional principal component analysis. To set up an `mxFDA` object from spatial single cell imaging data, calculate spatial summary functions, and perform exploratory data analysis and visualization of these spatial summary functions, see `vignette("mx_fda")`. To perform functional regression on spatial summary functions from multiplex imaging data, see  `vignette("mx_funreg")`.


```{r setup}
library(mxfda)
library(tidyverse)
library(patchwork)
```


# Functional data notation

The basic unit of observation is the curve $X_i(r)$ for subjects $i \in \ldots, I$ in the cross-sectional setting and $X_{ij}(r)$ for subject $i$ at sample $j \in \ldots, J_i$ for the multilevel structure, which occurs when there are multiple samples for each subject. Methods for functional data are typically presented in terms of continuous functions, but in practice data are observed on a discrete grid, $t$. In the `{mxfda}` package, $X_{ij}(r)$ is the spatial summary function (for example, Ripley's K) value at radius $r$ for sample $j$ from subject $i$.

# Functional principal components analysis (FPCA)

## Background on FPCA


FPCA characterizes modes of variability by decomposing functional observations into population level basis functions and subject-specific scores. The basis functions have a clear interpretation, analogous to that of PCA: the first basis function explains the largest direction of variation, and each subsequent basis function describes less. The FPCA model is typically written


$$
X_i(r) = \mu(r) + \sum_{k=1}^{K} c_{ik}\psi_{k}(r) + \epsilon_i(r)
$$

where $\mu(r)$ is the population mean, $\psi_{k}(r)$ are a set of orthonormal population-level functional principal components, $c_{ik}$ are subject-specific scores with mean zero and variance $\lambda_k$, and $\epsilon_i(r)$ are residual curves. Estimated functional principal components $\widehat{\psi}_1(r), \widehat{\psi}_2(r), \ldots, \widehat{\psi}_{K}(r)$ and corresponding variances $\widehat{\lambda}_1 \geq \widehat{\lambda}_2 \geq \ldots \geq \widehat{\lambda}_K$ are obtained from a truncated Karhunen-Lo\`eve decomposition of the sample covariance $\widehat{\Sigma}(s,t) = \widehat{\mbox{Cov}}(Y_i(s), X_i(r))$. The truncation lag $K$ is often chosen so that the resulting approximation accounts for at least 95\% of observed variance (@wrobel2016interactive). 

Conceptually, functional principal components can be thought of as patterns in the data that explain the most variance, and scores $c_{ik}$ are weights that describe how much each functional principal component contributes to the shape of a given subject's spatial summary function.


## Implementing FPCA

### Load and visualize data

Here we load data from the [VectraPolarisData](https://bioconductor.org/packages/release/data/experiment/html/VectraPolarisData.html) ovarian cancer dataset, which contains tumor microarray data for 128 women with high-grade serous ovarian cancer, where univariate nearest-neighbor G-functions for immune cells have already been extracted. This data has been modified, but the original data and experiment are described in @steinhart2021spatial. Each subject has one image.

```{r}
data("ovarian_FDA")
ovarian_FDA
```

These functions are visualized below.

```{r}
plot(ovarian_FDA, y = "fundiff", what = "uni g", sampleID = "patient_id") +
  geom_hline(yintercept = 0, color = "red", linetype = 2) +
  theme_minimal() 
  
```

Values above the dotted red line are interpreted as having a higher probability of observing a neighboring immune cell at radius $r$ than would be expected under complete spatial randomness. Conversely, values below the dotted line exhibit lower probability of observing a neighboring immune cell than expected under CSR.

### Run and visualize FPCA

Below we run FPCA on these G-function curves, using the `run_fpca()` function, and store these results in the `mxFDAobject` called `ovarian_FDA`. The argument `metric` is used to specify which spatial summary function to estimate FPCA for- in this case "uni g" specifies the univariate G function, `r` denotes which variable specifies the radius, and `value` denotes which value are the values of the spatial summary function. The argument `pve` takes a number from 0 to 1, and specifies the percent variance explained by the FPCA decomposition. For example, if `pve= 0.98` is selected, the resulting FPCA will return the number of FPCs that explain 98% of variance in the data.

```{r}
ovarian_FDA <- run_fpca(ovarian_FDA, 
                        metric = "uni g", 
                        r = "r", 
                        value = "fundiff",
                        pve = .95)

```

Note that the summary of this object now shows the number of functional principal components (FPCs) that have been calculated:

```{r}
summary(ovarian_FDA)
```


The FPCA results are stored in the `functional_pca` slot of the  `ovarian_FDA` object. To access this slot directly and view the extracted summary functions, type: 


```{r, eval = FALSE}
ovarian_FDA@functional_pca
```

We can visualize the results of FPCA using the `plot()` function (see `?plot.mxFDA` for details). The output of `plot()` is a `{ggplot2}` object which can then be easily added to/manipulated as any ggplot plot would.  Below we plot the first two functional principal components from our FPCA analysis of the G summary functions.


```{r fpc_plots, fig.width = 10}

p1 = plot(ovarian_FDA, what = 'uni g fpca', pc_choice = 1)
p2 = plot(ovarian_FDA, what = 'uni g fpca', pc_choice = 2)

(p1 + p2)
```

The plots above show $\hat{\mu}(r)\pm \sqrt{\hat{\lambda}_1\hat{\psi_k}(t)}$ for the first and second FPC. The black line in each plot is the mean of the data- this shows that the peak in probability of observing a neighboring immune cell occurs, on average, at about radius $r = 12$. In these plots, the blue and red lines represent one standard deviation of the FPC added (blue) or subtracted (red) from the population mean. For example in the plot for FPC 1, we can interpret the pattern of this FPC as a shift up or down from the population mean. Subjects who have a high score, $c_{i1}$ for FPC 1 have, across radii, a larger probability of observing a neighboring immune cell than average in the dataset. This pattern explains 73.8% of variance in the data.


We can further explore the results of FPCA interactively using the `{refund.shiny}` package from @wrobel2016interactive. Execute the code below to launch a Shiny app with interactive plots.

```{r refund.shiny, eval = FALSE}
G_fpca = extract_fpca_object(ovarian_FDA,
                             what = "uni g fpca", r = "r")

library(refund.shiny)
plot_shiny(G_fpca)

```

# Multilevel functional principal components analysis (MFPCA)


## MFPCA Background 


Multilevel functional principal components analysis (MFPCA) extends the ideas of FPCA to functional data with a multilevel structure. In the case of multiplex imaging data, this structure arises from multiple samples or ROIs for each subject. MFPCA models the within-subject correlation induced by repeated measures as well as the between-subject correlation modeled by classic FPCA. This leads to a two-level FPC decomposition, where level 1 concerns subject-specific effects and level 2 concerns sample-specific effects. Population-level basis functions and subject-specific scores are calculated for both levels (@di2009multilevel, @cui2023fast). The MFPCA model is:

$$
X_{ij}(r) = \mu(r)  + \sum_{k_1=1}^{K_1} 	c_{ik}^{(1)}\psi_{k}^{(1)}(r) + \sum_{k_2=1}^{K_2}c^{(2)}_{ijk}\psi_{k}^{(2)}(r) + \epsilon_{ij}(r)
$$

\noindent 
where $\mu(r)$ is the population mean, $\psi_{k}^{(1)}(r)$ and $\psi_{k}^{(2)}(r)$ are the functional principal components for levels 1 and 2, respectively, and $c_{ik}^{(1)}$ and $c^{(2)}_{ijk}$ are the subject-specific and subject-sample-specific scores. $K_1$ and $K_2$ represent the number of FPCs for levels 1 and 2, respectively.


## Implementing MFPCA

For MFPCA we will use the non-small cell lung carcinoma data, since we need multiple samples per subject. Below we extract the univariate K functions from the lung data.

```{r}
data(lung_df)

clinical = lung_df %>%
  select(image_id, patient_id, patientImage_id, gender, age, survival_days, survival_status, stage) %>%
  distinct()

spatial = lung_df %>%
  select(-image_id, -gender, -age, -survival_days, -survival_status, -stage)

mxFDAobject = make_mxfda(metadata = clinical,
                         spatial = spatial,
                         subject_key = "patient_id",
                         sample_key = "patientImage_id"
                         )

mxFDAobject = extract_summary_functions(mxFDAobject, 
                                        extract_func = extract_univariate,
                                        summary_func = Kest,
                                        r_vec = seq(0, 100, by = 1),
                                        edge_correction = "iso",
                                        markvar = "immune",
                                        mark1 = "immune")
```


These K-functions are visualized below.

```{r}
plot(mxFDAobject, y = "fundiff", what = "uni k", sampleID = "patientImage_id") +
  geom_hline(yintercept = 0, color = "red", linetype = 2)
```



### Run and visualize MFPCA

Below we run MFPCA on these K-function curves, using the `run_mfpca()` function, and store these results in the `mxFDAobject` lung cancer data object. Other arguments are the same as the `run_fpca()` function. From the summary of the object we see that 2 level 1 FPCs and 2 level 2 FPCs were calculated.

```{r}
mxFDAobject <- run_mfpca(mxFDAobject, 
                         metric = "uni k", 
                         r = "r", 
                         value = "fundiff",
                         pve = .99)

mxFDAobject
```


We can also visualize the results of MFPCA using the `plot()` function (see `?plot.mxFDA` for details).  Below we plot the first FPC for level 1 and the first FPC for level 2.


```{r}
p = plot(mxFDAobject, what = 'uni k mfpca', level1 = 1, level2 = 1)

p[[1]] + p[[2]]

```

The left plot shows $\hat{\mu}(r)\pm \sqrt{\hat{\lambda}_1^{(1)}}\hat{\psi_1}^{(1)}(t)$, and indicates that the first level 1 FPC explains essentially a shift up or down from the population mean. Subjects who have a high score, $c_{i1}$ for FPC 1 have, across radii, have more spatial clustering of immune cells than average in the dataset. The right plot shows $\hat{\mu}(r)\pm \sqrt{\hat{\lambda}_1^{(2)}}\hat{\psi}_1^{(2)}(t)$. This plot shows the pattern explained by FPC 1 from level 2. Essentially, it is the main pattern of variation explain differences across samples within a given subjects and shows that within a subject, differences are primarily driven by some images having higher overall K-function values and some being lower. 



# References
