---
title: "mx_funreg"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{mx_funreg}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  comment = "#>",
  fig.width = 8
)
```

```{r setup}
library(mxfda)
library(tidyverse)
library(patchwork)
```


## A note on data formatting

Use of this package requires that data be in a specific format: a long-form data frame with variables `id`, `index`, and `value`, where the `value` column contains functional observations for all subjects, the `id` column identifies which observations belong to which subject, and `index` provides the grid (domain) over which the `value`s are observed.  

The variable `id` should be a unique identifier in that each id identifies a single subject. Since we assume there is only one curve per subject for this package, `id` uniquely identifies each curve as well. Other covariates can be included in the data as long as the variables `id`, `index`, and `value` are present.


## Ovarian cancer multiplex imaging data

Here we load data from the Ovarian cancer dataset that where univariate nearest-neighbor G-functions for immune cells have already been extracted. See the vignette `mxfda::mx_fda` for more details on extracting spatial summary functions.


```{r load_data}
data("ovarian_FDA")
```


Visualize functions below. Interpret.

```{r}
plot(ovarian_FDA, y = "fundiff", what = "uni g", sampleID = "patient_id") +
  geom_hline(yintercept = 0, color = "red", linetype = 2) +
  theme_minimal() 
  
```


## Functional regression models for survival outcomes

### Cox regression using functional principal components as predictors

Runs FPCA and appends scores that explain up to 99% variance

```{r}
ovarian_FDA <- run_fpca(ovarian_FDA, id = "patient_id", metric = "uni g", r = "r", value = "fundiff",
                        lightweight = TRUE,
                        pve = .99)

```

visualize

```{r, fig.height = 8}
Gdf_fpc = bind_cols(ovarian_FDA@Metadata,
                    ovarian_FDA@`Functional PCA`$Gest$score_df)
p1 = Gdf_fpc %>%
  ggplot(aes(fpc1, survival_time, color = factor(event))) +
  geom_point() + ggtitle("fpc1") +
  theme(legend.position = "none")

p2 = Gdf_fpc %>%
  ggplot(aes(fpc2, survival_time, color = factor(event))) +
  geom_point() + ggtitle("fpc2") +
  theme(legend.position = "none")

p3 = Gdf_fpc %>%
  ggplot(aes(fpc3, survival_time, color = factor(event))) +
  geom_point() + ggtitle("fpc3") +
  theme(legend.position = "bottom")

p4 = Gdf_fpc %>%
  ggplot(aes(fpc4, survival_time, color = factor(event))) +
  geom_point() + ggtitle("fpc4") +
  theme(legend.position = "bottom")

(p1 + p2)/(p3 + p4)
```



Use fpc scores in cox regression model

```{r}
library(survival)
phmod_fpc = coxph(Surv(survival_time, event) ~ fpc1 + fpc2 + fpc3 + fpc4 + age, 
              data = Gdf_fpc)

summary(phmod_fpc)
```


### linear and additive functional Cox regression models

Linear functional cox model takes the form: 


Using age and gender and stage as scalar covariates


#### LFCM

```{r}
ovarian_FDA = run_fcm(ovarian_FDA, model_name = "fit_lfcm",
                      form = "survival_time ~ age",
                      metric = "uni g", id = "patient_id", r = "r", value = "fundiff",
                      analysis_vars = c("age", "survival_time", "event"),
                      afcm = FALSE)
```

```{r}
class(ovarian_FDA@`Functional Cox`$Gest$fit_lfcm)
```


```{r}
summary(ovarian_FDA@`Functional Cox`$Gest$fit_lfcm)
```

#### AFCM

```{r}
ovarian_FDA <- run_fcm(ovarian_FDA, model_name = "fit_afcm", 
                       form = "survival_time ~ age", metric = "uni g",
                    id = "patient_id", r = "r", value = "fundiff",
                    analysis_vars = c("age", "survival_time", "event"),
                    afcm = TRUE)
```

```{r}
class(ovarian_FDA@`Functional Cox`$Gest$fit_afcm)
```


```{r}
summary(ovarian_FDA@`Functional Cox`$Gest$fit_afcm)
```

#### AFCM, quantile transformed

```{r}
ovarian_FDA <- run_fcm(ovarian_FDA, model_name = "fit_afcm_qt",
                       form = "survival_time ~ age", metric = 'uni g',
                    id = "patient_id", r = "r", value = "fundiff",
                    analysis_vars = c("age", "survival_time", "event"),
                    afcm = TRUE,
                    quantile_transform = TRUE)
```


```{r}
summary(ovarian_FDA@`Functional Cox`$Gest$fit_afcm_qt)
```

## Visualize functional Cox models



Visualize the models

```{r extract_estimates, fig.with = 12}
Gdf = full_join(ovarian_FDA@Metadata,
                ovarian_FDA@`Univariate Summaries`$Gest)

lfcm = extract_surface(Gdf, ovarian_FDA@`Functional Cox`$Gest$fit_lfcm, analysis_vars = c("age")) %>%
  ggplot(aes(r, sumfun, fill = value)) +
  geom_tile() + 
  labs(y = "G-function value", title = "LFCM") +
  scale_fill_gradientn(colours=c("blue3", "#FFFFFF", "#F67F7F", "#FF0000")) +
  theme_minimal() +
  theme(legend.position = "bottom")

afcm = extract_surface(Gdf, ovarian_FDA@`Functional Cox`$Gest$fit_afcm, analysis_vars = c("age")) %>%
  ggplot(aes(r, sumfun, fill = value)) +
  geom_tile() + 
  labs(y = "G-function value", title = "AFCM") +
  scale_fill_gradientn(colours=c("blue3", "#FFFFFF", "#F67F7F", "#FF0000")) +
  theme_minimal() +
  theme(legend.position = "bottom")


lfcm + afcm 
```
