---
title: "mx_fpca"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{mx_fda}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  figs.out = "../figures"
)
```


This vignette describes the `mxfda` package for analyzing single-cell multiplex imaging data using tools from functional data analysis. Analyses for this package are executed and stored using an S4 object of class `mxFDA`. This vignette outlines how to set up an `mxFDA` object from spatial single cell imaging data, how to calculate spatial summary functions, and exploratory data analysis and visualization of these spatial summary functions. Details on how to perform downstream analysis and feature extraction using functional principal components analysis can be found in the separate vignette `mx_fpca`. To perform functional regression on spatial summary functions from multiplex imaging data, see the `mx_funreg` vignette.



```{r setup}
library(mxfda)
library(tidyverse)
library(patchwork)
devtools::load_all()
```


# Functional data notation


 The basic unit of observation is the curve $Y_i(t)$ for subjects $i \in \ldots, I$ in the cross-sectional setting and $Y_{ij}(t)$ for subject $i$ at visit $j \in \ldots, J_i$ for the multilevel or longitudinal structure. Methods for functional data are typically presented in terms of continuous functions, but in practice data are observed on a discrete grid that may be sparse or dense at the subject level and that may be the same across subjects or irregular.



# Functional principal components analysis (FPCA)

## Background on FPCA


Background on FPCA

FPCA characterizes modes of variability by decomposing functional observations into population level basis functions and subject-specific scores \citep{ramsay2005}. The basis functions have a clear interpretation, analogous to that of PCA: the first basis function explains the largest direction of variation, and each subsequent basis function describes less. The FPCA model is typically written
\beqn
\label{eq:fpca_model}
	Y_i(t) = \mu(t) + \sum_{k=1}^{K} c_{ik}\psi_{k}(t) + \epsilon_i(t)
\eeqn
\noindent
where $\mu(t)$ is the population mean,  $\psi_{k}(t)$ are a set of orthonormal population-level basis functions, $c_{ik}$ are subject-specific scores with mean zero and variance $\lambda_k$, and $\epsilon_i(t)$ are residual curves. Estimated basis functions $\widehat{\psi}_1(t), \widehat{\psi}_2(t), \ldots, \widehat{\psi}_{K}(t)$ and corresponding variances $\widehat{\lambda}_1 \geq \widehat{\lambda}_2 \geq \ldots \geq \widehat{\lambda}_K$ are obtained from a truncated Karhunen-Lo\`eve decomposition of the sample covariance $\widehat{\Sigma}(s,t) = \widehat{\mbox{Cov}}(Y_i(s), Y_i(t))$. In practice, the covariance $\widehat{\Sigma}(s,t)$ is often smoothed using a bivariate smoother that omits entries on the main diagonal to avoid a ``nugget effect" attributable to measurement error, and scores are estimated in a mixed model framework \citep{yao2005}. The truncation lag $K$ is often chosen so that the resulting approximation accounts for at least 95\% of observed variance. 



## Implementing FPCA


### Load and visualize data

Here we load data from the Ovarian cancer dataset that where univariate nearest-neighbor G-functions for immune cells have already been extracted. See the vignette `mxfda::mx_fda` for more details on extracting spatial summary functions.


```{r}
data("ovarian_FDA")
```



These functions are visualized below.

```{r}
plot(ovarian_FDA, y = "fundiff", what = "uni g", sampleID = "patient_id") +
  geom_hline(yintercept = 0, color = "red", linetype = 2) +
  theme_minimal() 
  
```


### Run and visualize FPCA

```{r}
mxFDAobject_oneImage <- run_fpca(ovarian_FDA, id = "patient_id", metric = "uni g", r = "r", 
                                 value = "fundiff",
                                 pve = .95)

```

Explain each of these plots



```{r fpc_plots, fig.width = 10}

p1 = plot(mxFDAobject_oneImage, what = 'uni g fpca', pc_choice = 1)
p2 = plot(mxFDAobject_oneImage, what = 'uni g fpca', pc_choice = 2)
p3 = plot(mxFDAobject_oneImage, what = 'uni g fpca', pc_choice = 3)
p4 = plot(mxFDAobject_oneImage, what = 'uni g fpca', pc_choice = 4)

(p1 + p2)/(p3 + p4)
```


INTERPRET WHAT YOU LEARNED FROM FPCA- this is an important piece


Shout out to refund.shiny

```{r refund.shiny, eval = FALSE}
G_fpca = extract_fpca_object(mxFDAobject_oneImage,
                             what = "uni g",
                             id = "patient_id", r = "r", value = "fundiff")

library(refund.shiny)
plot_shiny(G_fpca)

```



# MFPCA


## MFPCA Background 

Background on mfpca

Multilevel functional principal components analysis (MFPCA) extends the ideas of FPCA to functional data with a multilevel structure.


Multilevel functional data are increasingly common in practice; in the case of our DTI example, this structure arises from multiple clinical visits made by each subject. MFPCA models the within-subject correlation induced by repeated measures as well as the between-subject correlation modeled by classic FPCA. This leads to a two-level FPC decomposition, where level 1 concerns subject-specific effects and level 2 concerns visit-specific effects. Population-level basis functions and subject-specific scores are calculated for both levels \citep{di2009, di2014}. The MFPCA model is:
\beqn
\label{eq:mpca_model}
	Y_{ij}(t) = \mu(t) + \eta_j(t) + \sum_{k_1=1}^{K_1} 	c_{ik}^{(1)}\psi_{k}^{(1)}(t) + \sum_{k_2=1}^{K_2}c^{(2)}_{ijk}\psi_{k}^{(2)}(t) + \epsilon_{ij}(t)
\eeqn

\noindent 
where $\mu(t)$ is the population mean,  $\eta_j(t)$ is the visit-specific shift from the overall mean, $\psi_{k}^{(1)}(t)$ and $\psi_{k}^{(2)}(t)$ are the eigenfunctions for levels 1 and 2, respectively, and $c_{ik}^{(1)}$ and $c^{(2)}_{ijk}$ are the subject-specific and subject-visit-specific scores. Often, visit-specific means $\eta_j(t)$ are not of interest and can be omitted from the model. Estimation for MFPCA extends the approach for FPCA: estimated between- and within-covariances $\widehat{\Sigma}^{(1)}(s,t) = \widehat{\mbox{Cov}}(Y_{ij}(s), Y_{ij'}(t))$ for $j \neq j'$ and $\widehat{\Sigma}^{(2)}(s,t) = \widehat{\mbox{Cov}}(Y_{ij}(s), Y_{ij}(t))$ are derived from the observed data, smoothed, and decomposed to obtain eigenfunctions and values. Given these objects, scores are estimated in a mixed-model framework.


## Implementing MFPCA

Explain that there are multiple curves per subject for MFPCA and that's why lung data is better for mfpca

### Extract spatial summary functions and visualize data

Here we extract the univariate K function from the lung data. 

```{r}
data(lung_df)
clinical = lung_df %>%
  select(image_id, patient_id, patientImage_id, gender, age, survival_days, survival_status, stage) %>%
  distinct()

spatial = lung_df %>%
  select(-image_id, -gender, -age, -survival_days, -survival_status, -stage)

mxFDAobject = make_mxfda(metadata = clinical,
                         spatial = spatial,
                         key = "patient_id")

mxFDAobject = extract_summary_functions(mxFDAobject, "patientImage_id",
                                        extract_func = extract_univariate,
                                        summary_func = Kest,
                                        r_vec = seq(0, 100, by = 1),
                                        edge_correction = "iso",
                                        markvar = "immune",
                                        mark1 = "immune")


```


Visualize

```{r}
plot(mxFDAobject, y = "fundiff", what = "uni k", sampleID = "patientImage_id") +
  geom_hline(yintercept = 0, color = "red", linetype = 2)
```

Do some summaries (how many total functions? How many subjects? Range of number of functions per subject?)

### Run and visualize MFPCA

Right now this takes the same id in both slots and splits it under the hood. need to change this behavior


```{r}
# see what happens when you use mfpca.sc as an engine
mfpca_obj <- run_mfpca(mxFDAobject, 
                       id = "id", 
                                  image_id = "image_id", 
                                  metric = "uni k", r = "r", 
                                  value = "fundiff",
                                  pve = .99,
                       lightweight = FALSE)
```


Explore FPCA object

```{r}




p = plot_mfpc(mfpca_obj$mfpc_object, 1, 3)

p[[1]] + p[[2]]

#library(refund.shiny)
#plot_shiny(mfpca)
```

Calculate variance of level 2 scores. Also, are the 

```{r}

```




ICC



# FDA Tips / Principles/ Things to know

* Sparse vs. even grids- mention these, and what types of analysis you can do with each
* Better to have 20+ measurements per curve- that means using a dense grid of radii
* Number of knots cannot exceed the number of datapoints per function

# References
