---
title: "mx_funreg"
author: "Julia Wrobel"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{mx_funreg}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  comment = "#>",
  fig.width = 8
)
```

```{r setup}
library(mxfda)
library(tidyverse)
library(patchwork)
devtools::load_all()
```


## A note on data formatting

Use of this package requires that data be in a specific format: a long-form data frame with variables `id`, `index`, and `value`, where the `value` column contains functional observations for all subjects, the `id` column identifies which observations belong to which subject, and `index` provides the grid (domain) over which the `value`s are observed.  

The variable `id` should be a unique identifier in that each id identifies a single subject. Since we assume there is only one curve per subject for this package, `id` uniquely identifies each curve as well. Other covariates can be included in the data as long as the variables `id`, `index`, and `value` are present.


## Ovarian cancer multiplex imaging data

Load and clean data- explain dataset

```{r load_data}
# load processed ovarian cancer data
load(url("https://github.com/julia-wrobel/MI_tutorial/raw/main/Data/ovarian.RDA"))

# clean data
ovarian_df = ovarian_df %>%
  # subset to only analyze tumor areas
  filter(tissue_category == "Tumor") %>%
  # provide more intuitive patient and image IDs
  mutate(patient_id = as.numeric(factor(sample_id))) %>%
  # define cell type 'immune', which groups all immune cells
  mutate(immune = ifelse(phenotype_cd19 == "CD19+" | phenotype_cd8 == "CD8+" |
              phenotype_cd3 == "CD3+" | phenotype_cd68 == "CD68+", "immune", "other"),
         phenotype = case_when(phenotype_cd19 == "CD19+" ~ "B-cell",
                               phenotype_cd68 == "CD68+" ~ "macrophage",
                               TRUE ~ "other")) %>%
  select(patient_id, cell_id, x, y, age = age_at_diagnosis, immune, survival_time, 
         event = death, contains("phenotype")) 

```




### Extract spatial summary functions

To do functional data analysis, we first need to extract spatial summary functions from our multiplex imaging data. Below we extract the nearest-neighbor G-function for immune cells. See the vignette `mxfda::mx_fda` for more details on extracting spatial summary functions.


```{r}
Gdf = extract_summary_functions(ovarian_df, "patient_id",
                                extract_func = extract_univariate,
                                summary_func = Gest,
                                r_vec = seq(0, 50, by = 1),
                              edge_correction = "rs",
                markvar = "immune",
                mark1 = "immune",
                analysis_vars = c("age", "survival_time", "event"))
```


Visualize functions below. Interpret.

```{r}
Gdf %>%
  ggplot(aes(r, fundiff)) +
  geom_line(aes(group = patient_id), alpha = 0.2) +
  geom_hline(yintercept = 0, color = "red", linetype = 2) +
  theme_minimal() 
  
```


## Functional regression models for survival outcomes

### Cox regression using functional principal components as predictors

Runs FPCA and appends scores that explain up to 99% variance

```{r}
Gdf_fpc <- run_fpca(Gdf, id = "patient_id", r = "r", value = "fundiff",
                   analysis_vars = c("age", "survival_time", "event"),
                   lightweight = TRUE,
                   pve = .99)$mxfundata

```

visualize

```{r, fig.height = 8}
p1 = Gdf_fpc %>%
  ggplot(aes(fpc1, survival_time, color = factor(event))) +
  geom_point() + ggtitle("fpc1") +
  theme(legend.position = "none")

p2 = Gdf_fpc %>%
  ggplot(aes(fpc2, survival_time, color = factor(event))) +
  geom_point() + ggtitle("fpc2") +
  theme(legend.position = "none")

p3 = Gdf_fpc %>%
  ggplot(aes(fpc3, survival_time, color = factor(event))) +
  geom_point() + ggtitle("fpc3") +
  theme(legend.position = "bottom")

p4 = Gdf_fpc %>%
  ggplot(aes(fpc4, survival_time, color = factor(event))) +
  geom_point() + ggtitle("fpc4") +
  theme(legend.position = "bottom")

(p1 + p2)/(p3 + p4)
```



Use fpc scores in cox regression model

```{r}
library(survival)
phmod_fpc = coxph(Surv(survival_time, event) ~ fpc1 + fpc2 + fpc3 + fpc4 + age, 
              data = Gdf_fpc)

summary(phmod_fpc)
```


### linear and additive functional Cox regression models

Linear functional cox model takes the form: 


Using age and gender and stage as scalar covariates


#### LFCM

```{r}
fit_lfcm <- run_fcm(Gdf, form = "survival_time ~ age",
                    id = "patient_id", r = "r", value = "fundiff",
                    analysis_vars = c("age", "survival_time", "event"),
                    afcm = FALSE)
```

```{r}
class(fit_lfcm)
```


```{r}
summary(fit_lfcm)
```

#### AFCM

```{r}
fit_afcm <- run_fcm(Gdf, form = "survival_time ~ age",
                    id = "patient_id", r = "r", value = "fundiff",
                    analysis_vars = c("age", "survival_time", "event"),
                    afcm = TRUE)
```

```{r}
class(fit_afcm)
```


```{r}
summary(fit_afcm)
```

#### AFCM, quantile transformed

```{r}
fit_afcm_qt <- run_fcm(Gdf, form = "survival_time ~ age",
                    id = "patient_id", r = "r", value = "fundiff",
                    analysis_vars = c("age", "survival_time", "event"),
                    afcm = TRUE,
                    quantile_transform = TRUE)
```


```{r}
summary(fit_afcm_qt)
```

## Visualize functional Cox models



Visualize the models

```{r extract_estimates, fig.with = 12}
lfcm = extract_surface(Gdf, fit_lfcm, analysis_vars = c("age")) %>%
  ggplot(aes(r, sumfun, fill = value)) +
  geom_tile() + 
  labs(y = "G-function value", title = "LFCM") +
  scale_fill_gradientn(colours=c("blue3", "#FFFFFF", "#F67F7F", "#FF0000")) +
  theme_minimal() +
  theme(legend.position = "bottom")

afcm = extract_surface(Gdf, fit_afcm, analysis_vars = c("age")) %>%
  ggplot(aes(r, sumfun, fill = value)) +
  geom_tile() + 
  labs(y = "G-function value", title = "AFCM") +
  scale_fill_gradientn(colours=c("blue3", "#FFFFFF", "#F67F7F", "#FF0000")) +
  theme_minimal() +
  theme(legend.position = "bottom")

afcm_qt = extract_surface(Gdf, fit_afcm_qt, analysis_vars = c("age")) %>%
  ggplot(aes(r, sumfun, fill = value)) +
  geom_tile() + 
  labs(y = "G-function value", title = "AFCM, quantile transformed") +
  scale_fill_gradientn(colours=c("blue3", "#FFFFFF", "#F67F7F", "#FF0000")) +
  theme_minimal() +
  theme(legend.position = "bottom")


lfcm + afcm + afcm_qt
```

