---
title: "mx Functional Regression"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{mx Functional Regression}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  comment = "#>",
  fig.width = 8
)
invisible(suppressPackageStartupMessages(library(tidyverse)))
```

```{r setup}
#library(mxfda)
library(tidyverse)
library(patchwork)
devtools::load_all()
```


## A note on data formatting

Use of this package requires that data be in a specific format: a long-form data frame with variables `id`, `index`, and `value`, where the `value` column contains functional observations for all subjects, the `id` column identifies which observations belong to which subject, and `index` provides the grid (domain) over which the `value`s are observed.  

The variable `id` should be a unique identifier in that each id identifies a single subject. Since we assume there is only one curve per subject for this package, `id` uniquely identifies each curve as well. Other covariates can be included in the data as long as the variables `id`, `index`, and `value` are present.


## Ovarian cancer multiplex imaging data

Here we load data from the Ovarian cancer dataset that where univariate nearest-neighbor G-functions for immune cells have already been extracted. See the vignette `mxfda::mx_fda` for more details on extracting spatial summary functions.


```{r load_data}
data("ovarian_FDA")
```


Visualize functions below. Interpret.

```{r}
plot(ovarian_FDA, y = "fundiff", what = "uni g") +
  geom_hline(yintercept = 0, color = "red", linetype = 2) +
  theme_minimal() 
  
```


## Functional regression models for survival outcomes

### Cox regression using functional principal components as predictors

Runs FPCA and appends scores that explain up to 99% variance

```{r}
ovarian_FDA <- run_fpca(ovarian_FDA, metric = "uni g", r = "r", value = "fundiff",
                        lightweight = TRUE,
                        pve = .99)
ovarian_FDA
```

visualize

```{r, fig.height = 8}
Gdf_fpc = extract_fpca_scores(ovarian_FDA, 'uni g fpca')

p1 = Gdf_fpc %>%
  ggplot(aes(fpc1, survival_time, color = factor(event))) +
  geom_point() + ggtitle("fpc1") +
  theme(legend.position = "none")

p2 = Gdf_fpc %>%
  ggplot(aes(fpc2, survival_time, color = factor(event))) +
  geom_point() + ggtitle("fpc2") +
  theme(legend.position = "none")

p3 = Gdf_fpc %>%
  ggplot(aes(fpc3, survival_time, color = factor(event))) +
  geom_point() + ggtitle("fpc3") +
  theme(legend.position = "bottom")

p4 = Gdf_fpc %>%
  ggplot(aes(fpc4, survival_time, color = factor(event))) +
  geom_point() + ggtitle("fpc4") +
  theme(legend.position = "bottom")

(p1 + p2)/(p3 + p4)
```



Use fpc scores in cox regression model

```{r}
library(survival)
phmod_fpc = coxph(Surv(survival_time, event) ~ fpc1 + fpc2 + fpc3 + fpc4 + age, 
              data = Gdf_fpc)

summary(phmod_fpc)
```


### linear and additive functional Cox regression models

Linear functional cox model takes the form: 


Using age and gender and stage as scalar covariates


#### LFCM

```{r}
ovarian_FDA = run_fcm(ovarian_FDA, model_name = "fit_lfcm",
                      formula = survival_time ~ age, event = "event",
                      metric = "uni g", r = "r", value = "fundiff",
                      analysis_vars = c("age", "survival_time"),
                      afcm = FALSE)
```

```{r}
class(extract_model(ovarian_FDA, 'uni g', 'fit_lfcm'))
```


```{r}
summary(extract_model(ovarian_FDA, 'uni g', 'fit_lfcm'))
```

Visualize coefficient function

```{r}
lfcm_surface = extract_surface(ovarian_FDA, metric = "uni g", model = "fit_lfcm", analysis_vars = c("age"))
plot(lfcm_surface)
```



#### AFCM

```{r}
ovarian_FDA <- run_fcm(ovarian_FDA, model_name = "fit_afcm", 
                       formula = survival_time ~ age, event = "event",
                       metric = "uni g", r = "r", value = "fundiff",
                       analysis_vars = c("age", "survival_time"),
                       afcm = TRUE)
```

```{r}
class(extract_model(ovarian_FDA, 'uni g', 'fit_afcm'))
```


```{r}
summary(extract_model(ovarian_FDA, 'uni g', 'fit_afcm'))
```


Visualize the afcm model

```{r extract_estimates, fig.with = 12}
afcm_surface = extract_surface(ovarian_FDA, metric = "uni g", model = "fit_afcm", analysis_vars = c("age"), p = 0.05)
plot(afcm_surface)

```

## Model summaries and C-index 

C-index is a good way to compare across models, especially when incorporating cross validation.

```{r}
# calculate overall p value for fpca
#fit_fpc_null = coxph(Surv(survival_time, event) ~  age, data = Gdf_fpc)
#p_fpc = anova(fit_fpc_null, phmod_fpc)

fit_afcm = extract_model(ovarian_FDA, 'uni g', 'fit_afcm')
fit_lfcm = extract_model(ovarian_FDA, 'uni g', 'fit_lfcm')

c_index = c(
  rep(phmod_fpc$concordance[["concordance"]], times = 4),
  extract_c(fit_lfcm, Gdf_fpc$survival_time, Gdf_fpc$event),
  extract_c(fit_afcm, Gdf_fpc$survival_time, Gdf_fpc$event)
)


est_fpca = broom::tidy(phmod_fpc, exp = TRUE) %>%
  mutate(model = "fpca", edf = NA) %>%
  select(model,  term, "hazard ratio" = estimate, edf, p = p.value)

est_lfcm = bind_rows(broom::tidy(fit_lfcm) %>% 
  mutate("hazard ratio" = NA, p = NA, model = "lfcm") %>%
  select(model, term, "hazard ratio", edf, p),
  broom::tidy(fit_lfcm, parametric = TRUE, exp = TRUE) %>% 
  mutate(edf = NA, model = "lfcm") %>%
  select(model, term, "hazard ratio" = estimate, edf, p = p.value)
)

est_afcm = bind_rows(broom::tidy(fit_afcm) %>% 
  mutate("hazard ratio" = NA, p = NA, model = "afcm") %>%
  select(model, term, "hazard ratio", edf, p),
  broom::tidy(fit_afcm, parametric = TRUE, exp = TRUE) %>% 
  mutate(edf = NA, model = "afcm") %>%
  select(model, term, "hazard ratio" = estimate, edf, p = p.value)
)

bind_rows(est_fpca, est_lfcm, est_afcm) %>%
  filter(term != "age") %>%
  mutate(p = format.pval(p, digits = 1),
         "c-index" = c_index) %>%
  knitr::kable(digits = 2)
```
