---
title: "mx_funreg"
author: "Julia Wrobel"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{mx_funreg}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  comment = "#>",
  fig.width = 8
)
```

```{r setup}
library(mxfda)
library(tidyverse)
library(patchwork)
devtools::load_all()
```


## A note on data formatting

Use of this package requires that data be in a specific format: a long-form data frame with variables `id`, `index`, and `value`, where the `value` column contains functional observations for all subjects, the `id` column identifies which observations belong to which subject, and `index` provides the grid (domain) over which the `value`s are observed.  

The variable `id` should be a unique identifier in that each id identifies a single subject. Since we assume there is only one curve per subject for this package, `id` uniquely identifies each curve as well. Other covariates can be included in the data as long as the variables `id`, `index`, and `value` are present.


## Ovarian cancer multiplex imaging data

Load and clean data- explain dataset

```{r load_data}
data("ovarian_gfun")
```



Visualize functions below. Interpret.

```{r}
ovarian_gfun %>%
  ggplot(aes(r, fundiff)) +
  geom_line(aes(group = patient_id), alpha = 0.2) +
  geom_hline(yintercept = 0, color = "red", linetype = 2) +
  theme_minimal() 
  
```



## Functional regression models for survival outcomes

### Cox regression using functional principal components as predictors

Runs FPCA and appends scores that explain up to 99% variance

```{r}
Gdf_fpc <- run_fpca(Gdf, id = "patient_id", r = "r", value = "fundiff",
                   analysis_vars = c("age", "survival_time", "event"),
                   lightweight = TRUE,
                   pve = .99)$mxfundata

```

visualize

```{r, fig.height = 8}
p1 = Gdf_fpc %>%
  ggplot(aes(fpc1, survival_time, color = factor(event))) +
  geom_point() + ggtitle("fpc1") +
  theme(legend.position = "none")

p2 = Gdf_fpc %>%
  ggplot(aes(fpc2, survival_time, color = factor(event))) +
  geom_point() + ggtitle("fpc2") +
  theme(legend.position = "none")

p3 = Gdf_fpc %>%
  ggplot(aes(fpc3, survival_time, color = factor(event))) +
  geom_point() + ggtitle("fpc3") +
  theme(legend.position = "bottom")

p4 = Gdf_fpc %>%
  ggplot(aes(fpc4, survival_time, color = factor(event))) +
  geom_point() + ggtitle("fpc4") +
  theme(legend.position = "bottom")

(p1 + p2)/(p3 + p4)
```



Use fpc scores in cox regression model

```{r}
library(survival)
phmod_fpc = coxph(Surv(survival_time, event) ~ fpc1 + fpc2 + fpc3 + fpc4 + age, 
              data = Gdf_fpc)

summary(phmod_fpc)
```


### linear and additive functional Cox regression models

Linear functional cox model takes the form: 


Using age and gender and stage as scalar covariates


#### LFCM

```{r}
fit_lfcm <- run_fcm(ovarian_gfun, form = "survival_time ~ age",
                    id = "patient_id", r = "r", value = "fundiff",
                    analysis_vars = c("age", "survival_time", "event"),
                    afcm = FALSE)
```

```{r}
class(fit_lfcm)
```


```{r}
summary(fit_lfcm)
```

Visualize

```{r}
extract_surface(ovarian_gfun, fit_lfcm, analysis_vars = c("age"), se = TRUE) %>%
  ggplot(aes(r, exp(beta1))) +
  geom_line()+
  geom_line(aes(x=r, y = exp(beta1-2*beta1_se)), linetype = 'longdash')+
  geom_line(aes(x=r, y = exp(beta1+2*beta1_se)), linetype = 'longdash')+
  geom_hline(yintercept = 1, color = "red", linetype = 3) +
  labs(y = expression(e^hat(beta)(r)), title = "LFCM Hazard Ratio") +
  theme(axis.title = element_text(size = 15))
```


#### AFCM

```{r}
fit_afcm <- run_fcm(ovarian_gfun, form = "survival_time ~ age",
                    id = "patient_id", r = "r", value = "fundiff",
                    analysis_vars = c("age", "survival_time", "event"),
                    afcm = TRUE)
```

```{r}
class(fit_afcm)
```


```{r}
summary(fit_afcm)
```

Visualize

```{r}
bind_rows(extract_surface(ovarian_gfun, fit_afcm, analysis_vars = c("age")) %>% 
            mutate(type = "AFCM surface"),
          extract_surface(ovarian_gfun, fit_afcm, analysis_vars = c("age"), se = TRUE) %>% mutate(type = "p < 0.05")) %>%
  mutate(hr = exp(value)) %>%
    ggplot(aes(r, sumfun, fill = hr)) +
  geom_tile() + 
  labs(y = "G value") +
  scale_fill_distiller(name = expression(e^hat(F)(.,.)), na.value = "transparent",
                     type = "div", breaks = c(.9, .95, 1, 1.05, 1.1),
                     labels = c(".9"," .95", "1", "1.05", "1.1")) +
  facet_wrap(~type, ncol = 2) +
  theme(strip.text = element_text(size = 16))


```


```{r, eval = FALSE}
afcm = extract_surface(ovarian_gfun, fit_afcm, analysis_vars = c("age")) %>%
  ggplot(aes(r, sumfun, fill = value)) +
  geom_tile() + 
  labs(y = "G-function value", title = "AFCM") +
  scale_fill_gradientn(colours=c("blue3", "#FFFFFF", "#F67F7F", "#FF0000")) +
  theme_minimal() +
  theme(legend.position = "bottom")

```


## Model summaries and C-index 


```{r}
# calculate overall p value for fpca
fit_fpc_null = coxph(Surv(survival_time, event) ~  age, data = Gdf_fpc)
p_fpc = anova(fit_fpc_null, fit_fpc)


c_index = c(
  rep(fit_fpc$concordance[["concordance"]], times = 4),
  extract_c(fit_lfcm, Gdf_fpc$survival_time, Gdf_fpc$event),
  extract_c(fit_afcm, Gdf_fpc$survival_time, Gdf_fpc$event)
)


bind_rows(est_fpca, est_lfcm, est_afcm) %>%
  filter(term != "age") %>%
  mutate(p = format.pval(p, digits = 1),
         "c-index" = c_index) %>%
  knitr::kable(digits = 2)


```


